A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-21 paper.tex
minor intro edits

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This property is similar to 
  entropy  conservation
  following an adiabatic fluid.  We show that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  considered as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show in the limit of perfect coupling, that  dust settling in
itself cannot lead to axisymmetric instabilities, however thin the dust layer becomes.     
% 
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
buoyancy forces of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We discuss thermodynamical interpretations of 
dust-gas drag instabilities in disks, for example the streaming
instability, as a phase lag between the pressure 
and density of a dusty gas, which is   
analogous to pulsational instabilities 
in stars.  
\end{abstract}

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas drag introduces a 
wide range of new phenomena that are absent in pure gas
disks. Important examples include the streaming instability
\citep{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{chiang08,barranco09,lee10}. These lead to structure and
planetesimal formation, as well as turbulence. More recently, new
dusty instabilities have also been uncovered
\citep{loren15,loren16,lamb16}

%Understanding dusty gas dynamics is
%therefore essential to any planet formation theory. 
%phenomena that depends on gas drag 
%planet formation in dusty disks 
%dust component is building block of planets
%understanding dusty gas dynamics 
%small amount of dust can have significant consequences

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics with explicit
Lagragian dust particles \citep{nelson10,bai10,johansen11,yang14,zhu14,gibbons15,simon16,baruteau16}. In this
approach, the equations of motion for each particle is solved
including dust-gas drag, the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

Another approach is to model the dust population as a continous, pressureless
fluid
\citep{paardekooper06b,meheut12,laibe12,loren14,fu14b,surville16}. The equations
hydrodynamics are solved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$,
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of, and adapt, already-developed methods/codes for
simulating pure gas dynamics.

@@

It is possible to re-formulate the two-fluid equations into that for a
single fluid of for the total density $\rho$ and center-of-mass
velocity $\bm{v}$ of the dusty gas mixture
\citep{laibe14}. The standard hydrodynamic equations for
$\rho$ and $\bm{v}$ are supplemented by evolutionary equations for
the dust-to-gas ratio $\rhod/\rhog$ and the relative dust-gas velocity
$\Delta\bm{v}\equiv \bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$. The latter also contributes to 
additional source terms in the momentum and energy equations.  
The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can be numerically challenging for the two-fluid or Lagragian
particles approach; but is straight forward to model in the
one-fluid framework \citep{price15}.

The single fluid formalism also has the important advantage that it is
amenable to standard analytical methods
\citep{youdin05a,jacquet11}. This is the necessary  approach to gain physical insight.  
In this work, we are motivated to develop a general analytical
framework to  study the stability of dusty protoplanetary disks by
drawing on the physical and mathematical parallel between dusty gas and standard
hydrodynamics. This is possible in certain limits of astrophysical
interest.

We consider tightly coupled dust particles with $\tstop\Omega\ll
1$, where $\Omega$ is the disk orbital frequency. We may then use the
`terminal velocity approximation' to set $\Delta \bm{v} = \tstop\nabla
P/\rhog$, where $P$ is the gas pressure \citep{youdin05a,
  jacquet11,laibe14}. We also consider irradiated protoplanetary disks 
in which the temperature is set externally \citep{chiang97,stam08}. We
may then adopt a locally isothermal equation of state $P = c_s^2\rhog$
where $c_s$ is a prescribed sound-speed profile.  This physically
represents short cooling times \citep{lin15}.

Under these approximations, we show that the one-fluid equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas; and this property is analogous to the advection
of entropy following an adiabatic fluid. We will formally relate the
entropy of isothermal dusty gas to its dust content. Finite 
dust-gas drag can be re-interpreted as heating/cooling: the relative
motion between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

@@

By re-casting the dusty problem into identical form to 
hydrodynamics, we can apply standard methods to generalize previous stability 
results for pure gas disks to the dusty case. We demonstrate this for
perfectly coupled dust in strictly and locally isothermal disks. We
show that sharp dust edges may destabilize protoplanetary disks.  
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}. We provide 
thermodynamical interpretations of our dusty stability calculations.

This paper is organized as follows. In \S\ref{setup} we list the
equations describing tightly-coupled dusty gas. We show they are in 
fact the usual hydrodynamic equations. This leads to a precise 
definition of the entropy of an isothermal dusty gas. We describe approximate 
equilibrium solutions in \S\ref{eqm}. In \S\ref{limits} we discuss
general stability properties of the dusty problem in various
limits. We use these results to constraint the sharpness of dust edges
in isothermal disks. In \S\ref{linear_problem} we formulate an
explicit stability calcultion to study the dusty VSI, for which we
present numerical results in \S\ref{results}. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This property is similar to 
  entropy  conservation
  following an adiabatic fluid.  We show that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to generalize previous, well-known gaseous instabilities
  to dusty accretion disks.
%We show that strictly isothermal, non-self-gravitating, axisymmetric
%disks perfectly coupled to dust are generally stable, however thin the
%dust layer is. 
We show that in the limit of perfect coupling, dust settling in
itself cannot lead to axisymmetric instabilities, however thin the dust layer becomes.     
% 
We also extend 
the vertical shear instability (VSI) to locally isothermal, dusty 
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilize modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. 
These results can be understood in terms of 
buoyancy forces in an isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We discuss thermodynamical interpretations of 
dust-gas drag instabilities, for example the streaming
instability in protoplanetary disks, as a phase lag between the pressure 
and density of a dusty gas, which is   
analogous to pulsational instabilities 
in stars.  
\end{abstract}

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{chiang08,barranco09,lee10}. These lead to structure and
planetesimal formation, as well as turbulence. 
More recently, new 
dusty instabilities have also been uncovered through numerical
simulations \citep{loren15,loren16,lamb16}, but these are not yet well
understood.

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics coupled to explicit
Lagragian dust particles
\citep{nelson10,bai10,johansen11,yang14,zhu14,gibbons15,simon16,baruteau16}. 
The equation of motion for each particle is solved 
including dust-gas drag; the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

Another approach is to model the dust population as a continous, pressureless
fluid 
\citep{paardekooper06b,meheut12,laibe12,loren14,fu14b,surville16}. The 
hydrodynamic equations are evolved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$, 
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of, and adapt, already-developed numerical methods/codes for
simulating pure gas dynamics.

@@

It is possible to re-formulate the two-fluid equations into that for a
single fluid with total density $\rho$ and center-of-mass
velocity $\bm{v}$ of the dusty gas mixture
\citep{laibe14}. The standard hydrodynamic equations for
$\rho$ and $\bm{v}$ are supplemented by evolutionary equations for
the dust-to-gas ratio $\rhod/\rhog$ and the relative dust-gas velocity
$\Delta\bm{v}\equiv \bm{v}_\mathrm{d}-\bm{v}_\mathrm{g}$. The latter also contributes to 
additional source terms in the momentum and energy equations.

The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can be numerically challenging for the two-fluid and Lagragian
particles framework; but is straight forward to simulate in the
one-fluid formulation \citep{price15}.  The single fluid formalism
also has the important advantage that it is 
amenable to standard analytical methods
\citep{youdin05a,jacquet11}. This is the necessary  approach to gain
physical insight and explain some of the above newer dusty
instabilities.

In this work, we are motivated to develop a general analytical
framework to  study the stability of dusty protoplanetary
disks by drawing on the physical and mathematical parallel between dusty gas and standard
hydrodynamics. This is possible in certain limits of astrophysical
interest.

We consider tightly coupled dust particles with $\tstop\Omega\ll
1$, where $\Omega$ is the disk orbital frequency. We may then use the
`terminal velocity approximation' to set $\Delta \bm{v} = \tstop\nabla
P/\rhog$, where $P$ is the gas pressure \citep{youdin05a,
  jacquet11,laibe14}. %This eliminates the evolutionary equation for
%$\Delta \bm{v}$. 
 We also consider irradiated protoplanetary disks 
in which the temperature is set externally
\citep{chiang97,stam08}. This allows us to adopt a locally isothermal
equation of state, $P = c_s^2\rhog$, where $c_s$ is a prescribed
sound-speed profile. This physically represents short cooling times
\citep{lin15}.

Under these approximations, we show that the one-fluid equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas, which is is analogous to the advection
of entropy following an adiabatic fluid. We will formally relate the 
entropy of isothermal dusty gas to its dust content. Finite  
dust-gas drag can be interpreted as heating/cooling: the relative
motion between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

@@

By re-casting the dusty problem into identical form as standard  
hydrodynamics, we can apply well-known methods to generalize previous stability 
analyses for pure gas disks to the dusty case. We explicitly demonstrate this for
perfectly coupled dust in strictly and locally isothermal disks. We 
show that strictly isothermal protoplanetary disks are generally
stable to axisymmetric perturbations, regardless of how  thin the dust
layer is. We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}. We provide 
thermodynamical interpretations of our dusty stability calculations.

This paper is organized as follows. In \S\ref{setup} we list the
equations describing tightly-coupled dusty gas. We show they are in 
fact the usual hydrodynamic equations in disguise. 
We describe approximate  equilibrium solutions in \S\ref{eqm}. In \S\ref{limits} we discuss
general stability properties of the dusty disk in various
limits. There, we give a thermodynamical interpretation of
instabilities caused by directly by dust-gas drag. In
\S\ref{linear_problem} we list the linearized equations, which we
solve numerically in \S\ref{results} to study the dusty VSI. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.
